---
title: "Prepare Inputs for Qploidy"
date: "Last update: 2025-04-21"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette describes how to prepare the input files for use with `Qploidy`, an R package designed for ploidy, aneuploidy and large-scale copy number variation
 estimation using genotyping platform data. 

Before describing the inputs, it is important to understand the conditions under which the `Qploidy` methodology is effective, as well as 
its limitations. 
This understanding will help you determine whether `Qploidy` is the right tool for your dataset and research objectives.

# Prerequisites

### When does the `Qploidy` methodology work?

The `Qploidy` approach is effective under the following conditions:  
- Your marker data originates from **Axiom** or **Illumina genotyping arrays**.  
- Your marker data is derived from **targeted sequencing platforms** (e.g.,
  DArTag, GTseq, AgriSeq).  
- Your marker data is derived from **whole-genome sequencing** (make sure you have 
  sufficient coverage for accurate copy number estimation and you filter your data appropriately).
- All DNA samples were prepared following the **same library preparation
  protocol**.  
- You know the **ploidy** of at least a subset of 60 samples *or* you know the
  **most common ploidy** in the dataset.  
- Your dataset includes **heterozygous samples**.  

### When does the `Qploidy` methodology NOT work?

The methodology will not be effective under the following circumstances:  
- Your marker data comes from **RADseq** or **GBS** (Genotyping-by-Sequencing)
  platforms (sparse and inconsistent marker coverage across samples makes the signal standardization unreliable).  
- You intend to **combine datasets from different sequencing batches**.  
   - For example: If you extracted DNA and sequenced two plates (192 samples)
     as one batch, and later sequenced an additional three plates (288 samples)
     as a second batch, you would need to analyze the two batches **separately**
     in `Qploidy`. Combining all 480 samples into a single analysis will lead
     to incorrect results. See section "Check for batch effect" in the tutorial 
     "Aneuploid Detection in Alfalfa F1 Population" for more details.
- You **do not have a subset of samples with known ploidy** or **lack a
  predominant ploidy** in your dataset.  
    - Qploidy may still work depending on the quality of the data. Give it a try, but it is not guaranteed.
- Your samples consist exclusively of **inbred lines** (homozygous individuals), as dosage calling relies 
  on the segregation of allele intensities, which does not occur in fully homozygous lines.
- Your samples consist of first-generation, colchicine-treated individuals in which genome duplication was 
  directly induced, as the complete duplication of the genome doesn't generate all possible dosages of a polyploid. 
  Use progenies of these individuals instead, as they will segregate for all possible dosages.

# Installation

To install the `Qploidy` package, you can use the following commands:

```{r, eval=FALSE}
# Install the latest stable version from CRAN
install.packages("Qploidy")

# Install the development version from GitHub
#install.packages("devtools")
devtools::install_github("cristianetaniguti/Qploidy")
```

# Input files

`Qploidy` requires three main `data.frames` that we call:

- Data (`data`): contains the allele intensities or read counts for each marker and sample.
- Dosages (`genos`): contains the genotype dosages for each marker and sample of known ploidy or for all samples if overall ploidy is being used.
- Genomic positions (`genos.pos`): contains the genomic positions for each marker.

If you have standard format files, `Qploidy` provides functions to read them and convert them into these three required `data.frames`.

In this vignette, we demonstrate file format conversion using simulated data which mimics standard formats from:
- VCF (Variant Call Format) file
- Axiom array summary file
- Illumina array summary file

## `Data` data.frame format

### From VCF file

The required FORMAT in a VCF for Qploidy is the AD (Allele Depth) field, which 
contains the read counts for the reference and alternative alleles.
Let's generate our simulated VCF file. This file contains 500 markers and 60
samples, with a mix of tetraploid, diploid, and triploid samples. The tetraploid
samples are used as the reference for standardization because they are the most
common ploidy in the dataset. 

```{r}
library(Qploidy)

# Simulate a VCF file with 500 markers and 60 samples
simulate_vcf(
  seed = 1234, file_path = "vcf_example_simulated.vcf",
  n_tetraploid = 35, n_diploid = 5, n_triploid = 10,
  n_markers = 500
)
```

```{r}
data <- qploidy_read_vcf(vcf_file = "vcf_example_simulated.vcf")
head(data)
```

The resulting `data` data.frame contains the following columns: `MarkerName` and `SampleName` (identifiers), `X` and `Y` (allele signal 
intensities for the reference and alternative alleles, respectively), `R` (total signal intensity, i.e. `X + Y`), and `ratio` (the B-allele 
frequency, i.e. `Y / R`).


### From Axiom array summary file

In case your samples were genotyped using the Axiom array platform, you should
have a `summary` file or `.CEL` files. If you have `.CEL` files, you can convert
them into a `summary` file using the `apt-probeset-genotype` plugin available in
the Analysis Power Tools (APT) platform from Thermo Fisher Scientific (Waltham,
MA). A `summary` file is composed of a header that can have many lines and a body that 
looks like this:

```{r, echo=FALSE}
temp1 <- tempfile(fileext = ".txt")
simulate_axiom_summary(seed = 1234, file_path = temp1)
temp1 <- read.table(temp1, header = TRUE)
head(temp1)
```

In `Qploidy`, use the `read_axiom` function to read the `summary` file. It skips
the header and reads the body of the file. The function will return a data frame.

```{r, eval=FALSE}
data <- read_axiom("AxiomGT1_summary.txt")
head(data)
```

```
   MarkerName     SampleName         X         Y        R     ratio
1 AX-86752740      Lemon_Fiz  977.7548  793.4656 1771.220 0.4479768
2 AX-86752740   High_Voltage 1292.2205  582.7886 1875.009 0.3108191
3 AX-86752740    Lemon_Fiz.1  919.8055  870.9619 1790.767 0.4863624
4 AX-86752740 High_Voltage.1 1368.1266  645.9481 2014.075 0.3207170
5 AX-86752740     Brite_Eyes  238.5416 1242.7593 1481.301 0.8389648
6 AX-86752740 Morden_Blush.1  302.1575 1542.4037 1844.561 0.8361900
```

### From Illumina array 

In case you have a Illumina array summary file, `Qploidy` expects the
following format:

```
[Header]
GSGT Version    2.0.5
Processing Date 4/17/2023 9:31 AM
Content         specieX.bpm
Num SNPs        26251
Total SNPs      26251
Num Samples     100
Total Samples   100

[Data]
SNP Name   Sample ID   GC Score   Theta   X      Y      X Raw   Y Raw   Log R Ratio
mk-1       SAMP-2      0.6814     0.247   0.023  0.009  504     100     0.5211668
mk-10      SAMP-2      0.9204     0.000   0.006  0.000  346     27      -1.021583
mk-100     SAMP-2      0.1695     0.367   0.030  0.020  576     158     1.177404
mk-101     SAMP-2      0.8655     0.222   0.026  0.010  538     102     0.3459635
mk-103     SAMP-2      0.8932     0.000   0.019  0.000  465     18      0.7578704
mk-104     SAMP-2      0.5146     0.175   0.041  0.012  675     114     1.016254
mk-105     SAMP-2      0.8060     0.450   0.018  0.015  458     132     0.4305636
mk-106     SAMP-2      0.0000     0.386   0.002  0.002  315     57      NaN
mk-107     SAMP-2      0.4804     0.495   0.026  0.025  534     189     1.0151
```

Verify if your file looks like this one. If confirmed, you can use the `read_illumina_array`
function to read the data. If you include more than one file, the function will
merge them into a single data frame. 

```{r, eval=FALSE}
data <- read_illumina_array(
  "data/illumina/Set1.txt", # This is a example code, data not available
  "data/illumina/Set2.txt",
  "data/illumina/Set3.txt"
)
head(data)
```

```
MarkerName     SampleName     X     Y     R     ratio
1        CT1         S1-2_1 0.023 0.009 0.032 0.2812500
2       CT10         S1-2_1 0.006 0.000 0.006 0.0000000
3      CT100         S1-2_1 0.030 0.020 0.050 0.4000000
4      CT101         S1-2_1 0.026 0.010 0.036 0.2777778
5      CT103         S1-2_1 0.019 0.000 0.019 0.0000000
6      CT104         S1-2_1 0.041 0.012 0.053 0.2264151
```

## `Genos` data.frame format

The samples contained in the `genos` data frame will depend on the chosen path below.

### Define Reference Samples - Choose Path (1) or (2)

`Qploidy` applies a standardization method to allele intensities or read counts to enable 
the comparison of copy numbers across the genome of a sample. To achieve this, it requires 
a set of reference samples. There are two ways to select these reference samples:

1. **If you have a subset of samples with known ploidy**, your `genos` data.frame should only
contain the genotype dosages for this subset of samples. You will use this subset as the reference in the `Qploidy` standardization step.
2. **If you know the most common ploidy among your samples**, the `genos` data.frame should contain the genotype dosages for all samples. 
You can initially run `Qploidy` using all samples as the reference. From this first run, you can identify samples with the 
common ploidy, then select this subset to use as the reference in a second run of `Qploidy`. Check section `re-standardization` in the 
tutorial "Ploidy Estimated in a Mixed Population" for more details.

Below are the steps to proceed for each case, tagged as (1) and (2):

### (1) Using a Subset with Known Ploidy

For this method, you will need to separate your subset of samples with known ploidy:

In our simulated example, the sample names tell us which ploidy they are. Let's
create a vector with the tetraploid samples for us to use later:

```{r}
# All samples
all_samples <- unique(data$SampleName)
all_samples
tetraploid_samples <- all_samples[grep("Tetraploid", all_samples)]
tetraploid_samples
```

### (2) Using the Most Common Ploidy

For this method, you will use the entire dataset as the reference during the first 
round of the `Qploidy` run:

```{r, eval=FALSE}
data_reference <- data
```

### (1) and (2) Run Dosage Caller

To proceed, you need to determine the dosage for all reference samples. If you haven’t done 
this yet, you can use any suitable dosage caller.  

- For **array data**, we recommend using the 
[`fitPoly`]( https://CRAN.R-project.org/package=fitPoly) package.  
- For **sequencing data**, we suggest using the 
[`updog`](https://github.com/dcgerard/updog) or 
[`polyRAD`](https://github.com/lvclark/polyRAD) packages.  

These packages determine dosages by analyzing the distribution of allele intensities or read 
counts across all samples for each marker, using an informed ploidy model. You will provide 
either the **most common ploidy in your population** or the **known ploidy of your selected subset** 
as the informed ploidy.  

- **Path (2):** If you provide the most common ploidy, samples with different ploidy levels may 
receive incorrect dosages. However, these samples can be filtered out in the next step.  

**Warning:**  
Depending on the number of samples and markers, this step may take a significant amount of time to 
complete. It is highly recommended to run it on a high-performance computing system where you can 
utilize multiple cores and avoid issues with excessive RAM usage.

### VCF file

* If the VCF file already contains dosage/genotype (GT) information called for the subset/most 
common ploidy:

```{r}
genos <- qploidy_read_vcf("vcf_example_simulated.vcf", geno = TRUE)
dim(genos)

genos.pos <- qploidy_read_vcf("vcf_example_simulated.vcf", geno.pos = TRUE)
head(genos.pos)
```

If your VCF lacks the GT information for your reference samples, perform dosage 
calling using the AD field as described in the next section. After that, you can create 
the `genos` data frame with the dosage information and the `genos.pos` data frame with 
the genomic positions of the markers.

* If the VCF file lacks marker names in the `ID` column of the fixed portion, 
  the `MarkerName` column will contain `NAs`. To resolve this, you can use the 
  following code to concatenate the chromosome and position as marker names:

```{r, eval=FALSE}
genos.pos$MarkerName <- paste0(genos.pos$Chromosome, "_", genos.pos$Position)
head(genos.pos)
```


#### (1) Filter the genos object to keep only the known ploidy samples subset

```{r}
genos <- genos[which(genos$SampleName %in% tetraploid_samples), ]
```

* If the VCF file doesn't contain dosage/genotype information (GT format field) called 
for the subset/most common ploidy:

```{r}
library(tidyr)

# Prepare inputs for updog
## Approach (1)
# tobe_genotyped <- data[which(data$SampleName %in% tetraploid_samples),]

## Approach (2)
tobe_genotyped <- data

ref <- pivot_wider(tobe_genotyped[, 1:3], names_from = SampleName, values_from = X)
ref <- as.matrix(ref)
rownames_ref <- ref[, 1]
ref <- ref[, -1]
ref <- apply(ref, 2, as.numeric)
rownames(ref) <- rownames_ref

size <- pivot_wider(tobe_genotyped[, c(1, 2, 5)], names_from = SampleName, values_from = R)
size <- as.matrix(size)
rownames_size <- size[, 1]
size <- size[, -1]
size <- apply(size, 2, as.numeric)
rownames(size) <- rownames_size

size[1:10, 1:10]
ref[1:10, 1:10]
```

```{r}
library(updog)
multidog_obj <- multidog(
  refmat = ref,
  sizemat = size,
  model = "norm", # It depends on your population structure
  ploidy = 4, # Most common ploidy in your population
  nc = 2
) # Change the parameters accordingly!!

genos <- data.frame(
  MarkerName = multidog_obj$inddf$snp,
  SampleName = multidog_obj$inddf$ind,
  geno = multidog_obj$inddf$geno,
  prob = multidog_obj$inddf$maxpostprob
)

head(genos)
```

Notice that, if you are following approach (2), diploid and triploid sample genotypes will be incorrectly called as tetraploids. However, because 
they are not the most common samples in the dataset, they should not interfere significantly with the dosage calling of the tetraploids. You can
go back to this step after a first round of ploidy evaluation and call the dosages of the tetraploids only.

### Array data

Here we show an option of genotype calling for array data using fitpoly software:

```{r, eval=FALSE}
library(fitPoly)
saveMarkerModels(
  ploidy = 4, # Most common ploidy among Texas Garden Roses collection
  data = data_reference, # output of the previous function
  filePrefix = "fitpoly_output/texas_roses", # Define the path and prefix for the result files
  ncores = 2
) # Change it accordingly to your system!!!!

library(vroom) # Package to speed up reading files
fitpoly_scores <- vroom("fitpoly_output/texas_roses_scores.dat")

genos <- data.frame(
  MarkerName = fitpoly_scores$MarkerName,
  SampleName = fitpoly_scores$SampleName,
  geno = fitpoly_scores$maxgeno,
  prob = fitpoly_scores$maxP
)
head(genos)
```

```
   MarkerName  SampleName geno      prob
1 AX-86752740 1000 Wishes    2 0.7831438
2 AX-86752740  10004-N008    2 0.9892338
3 AX-86752740  10037_N046    2 0.9844171
4 AX-86752740  10038-N001    2 0.9916813
5 AX-86752740  10043_N019    2 0.9871057
6 AX-86752740  10043_N049    2 0.9955792
```

Unlike a VCF file, a `summary` file does not contain the genomic positions of the markers.
You will need to provide this information in a separate file, as shown in the following example.

```{r, eval=FALSE}
# File containing markers genomic positions
genos.pos_file <- read.table("geno.pos_roses_texas.txt", header = T)
head(genos.pos_file)
```

```
            SNP      probes chr      pos
1 Affx-86843634 AX-86752747   6   255852
2 Affx-86842821 AX-86752763   6 62032267
3 Affx-86839613 AX-86752769   5  9169310
4 Affx-86838724 AX-86752790   1 44259488
5 Affx-86840823 AX-86752809   7 12991931
6 Affx-86842443 AX-86752817   2 53861719
```

```{r, eval=FALSE}
# Edit for Qploidy input format
genos.pos <- data.frame(
  MarkerName = genos.pos_file$probes,
  Chromosome = genos.pos_file$chr,
  Position = genos.pos_file$pos
)
head(genos.pos)
```

```
   MarkerName Chromosome Position
1 AX-86752747          6   255852
2 AX-86752763          6 62032267
3 AX-86752769          5  9169310
4 AX-86752790          1 44259488
5 AX-86752809          7 12991931
6 AX-86752817          2 53861719
```

By the end of this tutorial you should have the three required data frames (`data`, `genos`, and `genos.pos`) ready to be used 
as input for `Qploidy`. You are now ready to run the analysis. Choose the tutorial that best fit your purpose to go next:

- [Aneuploid Detection with DArTag Markers in Alfalfa F1 Population](link_to_tutorial)
- [Copy Number Estimation in a Mixed Ploidy Population of Saccharomices sp. using WGS data](link_to_tutorial)
- [Large-scale Copy Number Variation Identification in a F1 Garden Roses Population using Axiom Array Data](link_to_tutorial)

# How to cite

### Version 1.0.0 - Standardization 

Taniguti, C. H., Lau, J., Hochhaus, T., Arias, D. C. L., Hokanson, S. C., Zlesak, D. C., Byrne, D. H.,
 Klein, P. E., & Riera-Lizarazu, O. (2025). Exploring chromosomal variations in garden roses: Insights 
 from high-density SNP array data and a new tool, Qploidy. The Plant Genome, e70044. 
 https://doi.org/10.1002/tpg2.70044

### Versions > 1.0.0 - HMM and BAF model selection

Manuscript in preparation. Please contact the author for more information.

# Acknowledgments

### Version 1.0.0 - Standardization 

Funded in part by the Robert E. Basye Endowment in Rose Genetics, Dept. of Horticultural Sciences, 
Texas A&M University, and USDA’s National Institute of Food and Agriculture (NIFA), Specialty Crop 
Research Initiative (SCRI) projects: ‘‘Tools for Genomics-Assisted Breeding in Polyploids: Development 
of a Community Resource’’ (Award No. 2020-51181-32156); and ‘‘Developing Sustainable Rose Landscapes via 
Rose Rosette Disease Education, Socioeconomic Assessments, and Breeding RRD-Resistant Roses with Stable 
Black Spot Resistance’’ (Award No. 2022-51181-38330).

### Versions > 1.0.0 - HMM and BAF model selection

Supported by [Breeding Insight](https://www.breedinginsight.org/).
