---
title: "Qploidy2 Tutorial"
subtitle: "Large-scale copy number estimation in alfalfa mapping population using DArTag genotyping data"
author: "Breeding Insight"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

This tutorial demonstrates a typical workflow for ploidy and aneuploidy estimation using the Qploidy2 R package. 
We will cover data import, standardization, model selection, Hidden Markov Model (HMM)-based copy number estimation, visualization, and VCF export. 

The dataset used here for our demonstration is an alfalfa mapping population genotyped using DArTag technology.
Refer to [this publication](https://doi.org/10.5061/dryad.pc866t1vd) for more information about the dataset.

To start, download the example dataset to your local directory:

```{r}
vcf_path_web <- "https://github.com/Breeding-Insight/BIGapp-PanelHub/raw/refs/heads/long_seq/alfalfa/GenoBrew_example/alfalfa_F1_marker_panel_dataset_publicly_available.vcf.gz"
download.file(vcf_path_web, destfile = "alfalfa_F1_marker_panel_dataset_publicly_available.vcf.gz")
vcf_path_local <- "alfalfa_F1_marker_panel_dataset_publicly_available.vcf.gz"
```

# Remarks

The Qploidy standardization methodology is effective under the following conditions:

* Your marker data originates from Axiom or Illumina genotyping arrays
* Your marker data is derived from targeted sequencing platforms (e.g., DArTag, GTseq, AgriSeq)
* Your marker data is derived from whole-genome sequencing platforms 
* All DNA samples were prepared following the same library preparation protocol (check [this section](#check-for-batch-effect-before-standardization) for more details)
For example: If you extracted DNA and sequenced two plates (192 samples) as one batch, 
and later sequenced an additional three plates (288 samples) as a second batch, you would need 
to analyze the two batches separately in Qploidy. Combining all 480 samples into a single analysis will lead to incorrect results due to batch effects.
* You know the ploidy of at least a subset (rough minimum of 60 samples) or you know the most common ploidy in the dataset.
The higher the number of samples with known or common ploidy, the better the results
* Your dataset includes heterozygous samples

The Qploidy2 B-allele frequency (BAF) model selection and HMM methodologies are effective under the following conditions:

* The ratio or BAF distributions of your samples present clear peaks corresponding to different copy numbers
* The total read depth or z-scores variations within a sample follows a normal distribution, with the mean shifting according to the copy number
* The sample includes heterozygous loci, which are essential for BAF-based ploidy estimation. 
Samples with heterozygosity lower than 5% will be considered haploid (CN = 1) and they cannot be distinguished from inbred lines with higher CN
* If allele counts are based on a sequencing platform, the sequencing should have enough depth to capture dosages of the 
highest possible copy number in the dataset
* The number of markers is sufficient to provide a good resolution for the BAF distributions
* The larger the amount of markers, the smaller the copy number variations that can be detected

# Install

* It requires R version superior to 3.6.0

Install Qploidy2 (if not already installed):

```{r, eval=FALSE}
devtools::install_github("Breeding-Insight/Qploidy2")
```

# Input data

`Qploidy2` receives three main objects named: 
- marker-level data (`data`): should contain sample names, marker ID, allele counts/intensities ratio and total for all samples
- marker positions (`geno.pos`): should contain marker IDs, and respective genomic chromosome and positions
- genotype dosages (`genos`) - only required for standardization: contain samples names, marker ID, 
and dosage/genotype considering expected ploidy. The samples in this file should be the ones with a known ploidy or the full dataset with 
dosage called considering the predominant ploidy. 

`Qploidy2` has functions to facilitate the conversion of standard file formats to these three objects. 
In this tutorial we will show the `qploidy_read_vcf` function for VCF conversion. The VCF must contain the AD field for allele counts.
Check function `?read_axiom` and `?read_illumina_array` if your data comes from genotyping arrays. 


```{r}
library(Qploidy2)

# Read marker-level data from VCF
data <- qploidy_read_vcf(vcf_path_local)
head(data)

# Read genotype dosages (optionally, restrict to samples with known ploidy)
genos <- qploidy_read_vcf(vcf_path_local, geno = TRUE)
head(genos)

# Read marker positions
geno.pos <- qploidy_read_vcf(vcf_path_local, geno.pos = TRUE)
head(geno.pos)
```

# Check data initial status

The `plot_raw` function provides diagnostic plots for metrics that will be considered by `Qploidy2` copy number estimation. 
Here is a brief description of each from top to bottom:

- Heterozygosity: y-axis = proportion of heterozygous loci in a genomic window with size defined by argument `window_size` (default = 2Mb);
x-axis = genomic positions
- Allele counts/intensities ratio (B/(A+B)): y-axis = ratio value; x-axis = genomic positions
- Ratio histogram split by chromosome: y-axis = observation counts; x-axis = ratio value
- Ratio histogram considering all markers of the samples: y-axis = observation counts; x-axis = ratio value
- Read depth (sequencing platforms) or total intensities (array) in y-axis and genomic position in x-axis

```{r, fig.width=11, fig.height=10, out.width="100%"}
samples <- unique(data$SampleName)
plot_raw(data = data, 
         geno.pos = geno.pos,
         sample = samples[1])
```

Check the plots for a few samples. If the histograms do not show clear peaks, consider applying `Qploidy` standardization.
If you already have clear peaks and do not want to try to improve them, skip to item [`# BAF Model Selection`](#BAF-Model-Selection).

## Check for batch effect before standardization

Qploidy2 standardization assumes that each marker behaves consistently across samples, while allowing markers to differ from one another. 
Depending on the genotyping platform, however, this assumption may not hold — particularly if DNA extraction protocols were modified or samples were processed in different plates or batches.

For example, you may have extracted DNA at one time point and sequenced a plate with 96 samples in a given year, then used the stored DNA two years later to sequence an additional 96 samples. 
Changes in DNA quality over time can affect probe performance, leading to systematic differences between batches. 
In such cases, standardization should be performed separately for each batch.

You can assess whether batch effects are present in your dataset using the following function:

```{r}
# provide a passport file (containing year of DNA extraction/plate number/ or any possible factor for a batch effect)
# This example dataset doesn't have a passport available, we will simulate one just to exemplify the function usage:
set.seed(123)
samples <- unique(data$SampleName)
passport_file <- data.frame(SampleName = samples, year_seq = sample(rep(c(2020, 2022, 2025), length(samples))))
head(passport_file)
```

```{r}
p <- pca_plot(input = data, 
              geno.pos = geno.pos, 
              passport_file = passport_file, 
              group_column = "year_seq", 
              sampleID_column = "SampleName",
              col2use = "R")

p$PC1xPC2

```

In this example, we did not detect evidence of batch effects, as samples did not cluster according to genotyping year. 
Therefore, we can proceed with standardization using the combined dataset.

## Initial screening for chromosome abnormalities (optional)

For some genotyping platforms, total read depth is consistent across samples. 
When this assumption holds for your dataset, deviations in a sample’s total depth relative to the population average can provide an initial indication of copy number variation.

You can rapidly screen for candidate samples using the function below:

```{r, fig.width=6, fig.height=5, out.width="100%"}
chromosome_level_test_plot_qploidy(input = data, 
                                   geno.pos = geno.pos, 
                                   selected_samples = samples[1:12], # plot the first 12 
                                   col2test = "R")
```

Before proceeding to the standardization step, consider removing abnormal chromosomes from your reference dataset (genos).

## Overview of samples heterozygosity

```{r}
# interactive graphic if plotly package is installed
p <- plot_heterozygosity(data, col2use = "ratio")
p
```

This is an interactive graphic; hover your mouse over the cells to get information on sample name, heterozygosity percentage, and the number of markers used for this estimation.
Heterozygosity is an important metric for Qploidy2, as it uses the allele ratio heterozygous peaks as part of the input data for estimating ploidy. Samples with heterozygosity lower than 5% will be considered haploid (CN = 1) and cannot be distinguished from inbred lines with higher CN.


# Standardization

Standardization of markers raw read counts or intensity data is essential for reducing technical noise 
and improving the accuracy of ploidy estimation. Markers from diverse platforms have different 
characteristics, and standardization helps to make them comparable. 

The function has several parameters that can be adjusted based on the characteristics of your dataset.
You can access a description of all parameters in `?standardize`, but here we will focus only on the most important ones:
`ploidy.standardization` and `threshold.n.clusters`:

**Parameters:**
- `genos`: This data.frame defines which samples are your reference for the standardization. If you have samples with known ploidy, filter this object to keep only them. You can use the command:

```{r, eval = FALSE}
# Demonstration code, not applied to current dataset
# known_tetraploids <- paste0("Sample",1:100)
# str(known_tetraploids)
# genos <- genos[which(genos$SampleName %in% known_tetraploids),]
```

If you are not sure about ploidy of at least 60 samples, just keep all them and define the most common ploidy as `ploidy.standardization`.

- `ploidy.standardization`: Sets the predominant ploidy for the dataset defined in the object `genos` (e.g., 2 for diploid, 4 for tetraploid). This guides clustering and normalization, and should match the known or most frequent ploidy in your samples for best results.

- `threshold.n.clusters`: Minimum number of genotype clusters required for a marker to be retained. The maximum is defined as the possible number of the dosages according to the ploidy
defined on `ploidy.standardization` (ploidy + 1). Lower values requires imputation of dosage clusters for missing genotypes. The imputation is performed based on distance to other clusters, 
so it is recommended to have at least 3 clusters for a good imputation. If too many markers are filtered, relax this threshold (e.g., to 4 for tetraploids) to retain more markers with imputed clusters.


```{r}
qploidy_standardization <- standardize(
  data = data,
  genos = genos,
  geno.pos = geno.pos,
  ploidy.standardization = 4, # Set to known or most frequent ploidy
  threshold.n.clusters = 5,   # Minimum clusters for marker retention
  n.cores = 2,
  out_filename = "standardization.tsv.gz",
  verbose = TRUE, 
)

# Check filtering summary
qploidy_standardization
```

If too many markers are filtered, relax the cluster threshold:

```{r}
qploidy_standardization <- standardize(
  data = data,
  genos = genos,
  geno.pos = geno.pos,
  ploidy.standardization = 4,
  threshold.n.clusters = 4,
  n.cores = 2,
  out_filename = "standardization.tsv.gz",
  verbose = TRUE
)
qploidy_standardization
```

You can reload a saved standardization object:

```{r, eval=FALSE}
qploidy_standardization <- read_qploidy_standardization("standardization.tsv.gz")
```

See examples on how the standardization process works for a few markers in the dataset:

```{r}
markers <- unique(qploidy_standardization$data$MarkerName)
six_markers <- qploidy_standardization$data[which(qploidy_standardization$data$MarkerName %in% markers[c(4,19,24,51,56,57)]),]
plot_geno_by_marker(six_markers, alpha = 0.1)
```

# Quality Assessment and Visualization

Visualize the effect of standardization for each sample and explore BAF and Z-score distributions:

```{r}
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = "J432",
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))

plot_qploidy_standardization(x = qploidy_standardization,
                             sample = "J432",
                             type = c("zscore", "R"))
```

See `?plot_qploidy_standardization` for more plot types.

## Chromosome-level screening for abnormalities based on Z-score (optional)

You can use the same quick check we used before on the raw total depth (R) with the standardized z-scores. This is useful to identify samples with potential aneuploidies or large CNVs before 
running the HMM, and also to select samples with known ploidy to guide the BAF model selection.

```{r, fig.width=6, fig.height=5, out.width="100%"}
chromosome_level_test_plot_qploidy(input = qploidy_standardization, 
                                   selected_samples = samples[1:12], 
                                   col2test = "z")
```

# BAF Model Selection

Select the best BAF model for a sample. The approach implemented here for this step is based on [`nQuack`](https://github.com/mgaynor1/nQuack)
methods. It tests different statistical distribution mixture models fitting for each tested ploidy on the BAF distributions and selects the best based on BIC.
The idea is the same, but the implementation here differs. 
`nQuack` has a robust EM algorithm able to return tailored parameters for the selected distribution to users. 
`Qploidy2` does not use an EM algorithm but a grid of parameters to be tested, 
which is less time- and computationally demanding but may not provide the best parameters for this step. 
This does not compromise the quality of `Qploidy2` final results since an EM is implemented together with the HMM in the next step of our workflow.
However, if you want to also have the best results in this step, we suggest running also `nQuack` to check the differences and provide this function with the best parameters.

This step helps determine the most likely ploidy and mixture 
model parameters for BAF. The selected model will be used in the HMM for copy number estimation. 
You can select a sample with known ploidy (e.g., from the previous step) to guide the model selection, 
or test multiple samples and compare results.

## Applying to raw data

```{r}
sample_ratios <- data$ratio[data$SampleName == "J432"]
selected_model_raw <- select_best_baf_model(baf_vec = sample_ratios,
                                        cn_grid= c(2,3,4),
                                        bw = c(0.02, 0.04,0.1),
                                        plot = TRUE)

selected_model_raw$plot
head(selected_model_raw$grid_results)
```

For this dataset, standardization is required for better results. This step and other downstream steps only make this move evident.

## Applying to standardized data

```{r}
selected_model <- select_best_baf_model(qploidy_standardization = qploidy_standardization, 
                                        sample = "J432",
                                        cn_grid= c(2,3,4),
                                        bw = c(0.02, 0.04,0.1),
                                        plot = TRUE)

selected_model$plot
head(selected_model$grid_results)
```

# HMM Copy Number Estimation (Single Sample)

Estimate copy number for a sample using a Hidden Markov Model. 
This step provides a robust approach that integrates BAF and Z-score (standardized total depth) 
information across genomic regions to infer CNV segments. Find all parameters description in `?hmm_estimate_CN`.

## Applying to raw data

```{r}
res_raw <- hmm_estimate_CN(data = data, 
                       geno.pos = geno.pos,
                       selected_model = selected_model_raw, 
                       use_values = c("ratio","R"),
                       sample_id = "J432",
                       cn_grid =  c(2,3,4,5,6))

# View results
res_raw
```

Visualize the CNV track:

```{r , fig.width=11, fig.height=10, out.width="100%"}
p <- plot_cn_track(hmm_CN = res_raw, data = data, 
                   geno.pos = geno.pos, 
                   sample_id = "J432",
                   show_window_lines = FALSE,
                   z_by_mean = TRUE)

# All panels together
p$arranged
```

## Applying to standardized data

```{r}
res <- hmm_estimate_CN(qploidy_standarize_result = qploidy_standardization,
                       selected_model = selected_model,
                       sample_id = "J432",
                       cn_grid =  c(2,3,4,5,6))

# View results
res
```

Visualize the CNV track:

```{r , fig.width=11, fig.height=10, out.width="100%"}
p <- plot_cn_track(hmm_CN = res,
                   qploidy_standarize_result = qploidy_standardization,
                   sample_id = "J432",
                   show_window_lines = FALSE,
                   z_by_mean = TRUE)

# All panels together
p$arranged
```

```{r}
# Individual panels
p$baf
p$z
p$cn
p$baf_dist
```

Summarize results:

```{r}
summ_hmm_chr <- summarize_cn_mode(res, level = "chromosome")
summ_hmm_sample <- summarize_cn_mode(res, level = "sample")
```


# HMM Copy Number Estimation (Multiple Samples)

Run HMM for all samples in parallel. The model will be selected internally using `select_best_baf_model`; you just need to define the grids (`bw`, `add_uniform_grid`, and `uniform_weight_grid`) according to what you observed when running it on a single sample.

**Note**: Running HMM for many samples can be computationally intensive. Adjust `n_cores` based on your resources and dataset size.

## Applying to raw data

```{r}
multi_esti_raw <- hmm_estimate_CN_multi(data = data,
                                    geno.pos = geno.pos,
                                    use_values = c("ratio","R"),
                                    sample_ids = "all",
                                    n_cores = 2, 
                                    cn_grid =  c(2,3,4,5,6),
                                    bw = c(0.02, 0.04,0.1))
```

Visualize CNV tracks for multiple samples:

```{r, fig.width=11, fig.height=10, out.width="100%"}
compare_cn_track(hmm_CN = multi_esti_raw,
                 samples_to_plot = samples[1:30],
                 facet_ncol = 8)
```

Again, not the best results for this dataset. Standardization is required.

## Applying to standardized data

```{r}
multi_esti <- hmm_estimate_CN_multi(qploidy_standarize_result = qploidy_standardization,
                                    sample_ids = "all",
                                    n_cores = 2, 
                                    cn_grid =  c(2,3,4,5,6),
                                    bw = c(0.02, 0.04,0.1))
```

Visualize CNV tracks for multiple samples:

```{r, fig.width=11, fig.height=10, out.width="100%"}
compare_cn_track(hmm_CN = multi_esti,
                 samples_to_plot = samples[31:50],
                 facet_ncol = 8)
```

```{r}
# For a specific sample
p <- plot_cn_track(hmm_CN = multi_esti,
                   qploidy_standarize_result = qploidy_standardization,
                   sample_id = "85-182",
                   show_window_lines = TRUE)
p$arranged
```

```{r}
p$baf
```

# Re-standardization (Optional)

If you want to improve BAF resolution (e.g., after identifying a subset of samples with known ploidy via HMM), re-standardize:

```{r}
re_qploidy_standardization <- re_standardize(
  data = data,
  geno.pos = geno.pos,
  hmm_CN_multi  = multi_esti,
  selected_model = selected_model,
  threshold.geno.prob = 0.9,
  ploidy.standardization = 4,
  threshold.n.clusters = 4,
  n.cores = 2,
  out_filename = "re_standardization.tsv.gz",
  verbose = TRUE
)

re_qploidy_standardization

# Compare before and after
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = "85-295",
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
plot_qploidy_standardization(x = re_qploidy_standardization,
                             sample = "85-295",
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
```

## Re-run the HMM with the new standardization

```{r}
re_multi_esti <- hmm_estimate_CN_multi(
  qploidy_standarize_result = re_qploidy_standardization,
  sample_ids = "all",
  n_cores = 2,
  cn_grid =  c(2,3,4,5,6),
  bw = c(0.02, 0.04,0.1), 
  add_uniform_grid = TRUE, 
  uniform_weight_grid = c(0.2, 0.5,0.95))

# CNV profiles before
compare_cn_track(hmm_CN = multi_esti,
                 samples_to_plot = samples[91:110],
                 facet_ncol = 8)

# CNV profiles after
compare_cn_track(hmm_CN = re_multi_esti,
                 samples_to_plot = samples[91:110],
                 facet_ncol = 8)
```

```{r, fig.width=11, fig.height=10, out.width="100%"}
# Same plots can be generated
p_re <- plot_cn_track(hmm_CN = re_multi_esti,
                      qploidy_standarize_result = re_qploidy_standardization,
                      sample_id = "85-295",
                      show_window_lines = TRUE)

p <- plot_cn_track(hmm_CN = multi_esti,
                   qploidy_standarize_result = qploidy_standardization,
                   sample_id = "85-295",
                   show_window_lines = TRUE)

p$arranged #before
p_re$arranged #after
```

For this dataset, there are a few changes for some samples like the one showed above. 
Chromosome 7 is pointed with higher number (5) of copies in the first round standardization and with 
the re-standardization it is presented as 4 copies. This instability still leaves doubts about 
the CNV of this particular chromosome. You can further explore tweaking the other parameters, 
for example, increasing the number of markers in the windows to better fit the mixed models on the BAF distributions:

```{r, fig.width=11, fig.height=10, out.width="100%"}

res_tweak <- hmm_estimate_CN(qploidy_standarize_result = re_qploidy_standardization,
                       selected_model = selected_model,
                       sample_id = "85-295",
                       min_snps_per_window = 60,
                       cn_grid =  c(2,3,4,5,6))

p <- plot_cn_track(hmm_CN = res_tweak,
                   qploidy_standarize_result = re_qploidy_standardization,
                   sample_id = "85-295",
                   show_window_lines = TRUE)

p$arranged
```

Explore more parameters consulting `?hmm_estimate_CN`.

Once you get to a conclusion with results better explain the patterns, you can update your multi-sample
HMM object by using the functions:

```{r}
final_res <- update_hmm_CN_multi(multi_esti, 
                                 res_tweak, 
                                 rm_sample = FALSE)
```

If you decide that the results for this sample is inconclusive, 
you can set the `rm_sample` parameter to `TRUE` to remove it from your final results.

# Call dosages

**Warning**: Beta version - Dosage call requires more testing and review.

Qploidy2 uses the BAF distributions to call dosages. See an example:

```{r}
# res was generated by hmm_estimate_CN above, it contains only one sample 
sample_name <- "85-17"
baf <- final_res$by_marker$baf[final_res$by_marker$SampleName == sample_name]

selected_model <- select_best_baf_model(baf_vec =  baf, 
                                        sample = sample_name,
                                        cn_grid= c(2,3,4),
                                        bw = c(0.02, 0.04,0.1),
                                        plot = TRUE)

dosages_one_sample <- call_BAF_dosages(baf, selected_model = selected_model, plot=TRUE)
head(dosages_one_sample$data)

dosages_one_sample$plot
```

Call dosages for all samples considering the copy number estimated:

```{r}
dosages <- call_hmm_dosages(hmm_CN = final_res, selected_model)
head(dosages)
```

# Export VCF

```{r}
export_VCF(dosages, file = "alfalfa_Qploidy_CN_dosages.vcf")
```

You can read the VCF with vcfR and extract fields:

```{r}
library(vcfR)
vcf <- read.vcfR("alfalfa_Qploidy_CN_dosages.vcf")
vcf@gt[1:5,1:5]

CN <- extract.gt(vcf, "CN")
CN[1:5,1:5]

GT <- extract.gt(vcf, "GT")
GT[1:5,1:5]

dosage_prob <- extract.gt(vcf, "PMD")
dosage_prob[1:5,1:5]
```

# Plotting all samples in batches (Optional)

Generate CNV plots for all samples in batches:

```{r, eval=FALSE}
n_plots_by_page <- c(seq(1, length(samples), by = 40), length(samples))
for(i in 1:(length(n_plots_by_page) -1)){
  comp_p <- compare_cn_track(hmm_CN = final_res,
                             samples_to_plot = samples[n_plots_by_page[[i]]:n_plots_by_page[i+1]],
                             facet_ncol = 12)
  ggsave(comp_p, filename = paste0("Samples_", n_plots_by_page[i], "_", n_plots_by_page[i +1], ".png"), height = 10, width = 11)
}
```


# How to cite

### Qploidy - Standardization 

Taniguti, C. H., Lau, J., Hochhaus, T., Arias, D. C. L., Hokanson, S. C., Zlesak, D. C., Byrne, D. H.,
Klein, P. E., & Riera-Lizarazu, O. (2025). Exploring chromosomal variations in garden roses: Insights 
from high-density SNP array data and a new tool, Qploidy. The Plant Genome, e70044. 
https://doi.org/10.1002/tpg2.70044

### Qploidy2 - HMM and BAF model selection

Manuscript in preparation. Please contact the author for more information.

# Acknowledgments

### Qploidy version 1.0.0 

Funded in part by the Robert E. Basye Endowment in Rose Genetics, Dept. of Horticultural Sciences, 
Texas A&M University, and USDA’s National Institute of Food and Agriculture (NIFA), Specialty Crop 
Research Initiative (SCRI) projects: ‘‘Tools for Genomics-Assisted Breeding in Polyploids: Development 
of a Community Resource’’ (Award No. 2020-51181-32156); and ‘‘Developing Sustainable Rose Landscapes via 
Rose Rosette Disease Education, Socioeconomic Assessments, and Breeding RRD-Resistant Roses with Stable 
Black Spot Resistance’’ (Award No. 2022-51181-38330).

### Qploidy versions > 1.0.0 and Qploidy2 

Supported by [Breeding Insight](https://www.breedinginsight.org/).
